\section{List-Question Quick Reference}
\label{sec:lists}

Memorizable lists for the ``enumerate'' question type. Each item has a brief
gloss; in the exam, give the list and a one-line gloss per item.

% ---------------------------------------------------------------------
\subsection{Raw-data issues in data tables}
\begin{itemize}[leftmargin=1.4em]
  \item \term{Missing values} -- empty cells / NA; need imputation or removal.
  \item \term{Errors / outliers} -- typos, sensor faults, impossible values.
  \item \term{Duplicates} -- repeated records inflate counts.
  \item \term{Inconsistent units / formats} -- km vs.\ miles, date formats.
  \item \term{Inconsistent encoding} -- ``M''/``Male''/``1'' for same category.
  \item \term{Wrong data types} -- numbers stored as text, categorical as int.
  \item \term{Mixed granularity} -- rows at different aggregation levels.
  \item \term{Non-tidy structure} -- values in column names, multiple variables
        per column.
  \item \term{Skew / heavy tails / scale differences} -- may need log or
        normalization.
  \item \term{Imbalanced classes} -- distorts colour/area perception.
\end{itemize}

% ---------------------------------------------------------------------
\subsection{Factors affecting scalability}
\begin{itemize}[leftmargin=1.4em]
  \item \term{Number of data items} (rows) -- overplotting, slow rendering.
  \item \term{Number of dimensions / attributes} -- curse of dimensionality,
        too many axes/panels.
  \item \term{Number of categories} -- exhausts distinguishable colours/shapes.
  \item \term{Display resolution / screen pixels} -- pixel budget limits marks.
  \item \term{Human perception} -- limited channels, pre-attentive capacity
        (\term{perceptual scalability}).
  \item \term{Rendering / computation budget} -- frame rate, memory.
  \item \term{Interaction latency} -- query/response time
        (\term{interactive scalability}).
\end{itemize}

% ---------------------------------------------------------------------
\subsection{Sources of latency}
\begin{itemize}[leftmargin=1.4em]
  \item \term{Data transfer / network} -- moving data to the client.
  \item \term{Querying / aggregation} -- on-the-fly computation over big data.
  \item \term{Rendering} -- drawing many marks each frame.
  \item \term{Layout computation} -- e.g.\ force-directed iterations, DR.
  \item \term{Disk / IO} -- loading from storage.
  \item \term{Interaction round-trip} -- client$\leftrightarrow$server latency
        on brush/filter.
\end{itemize}

% ---------------------------------------------------------------------
\subsection{Aesthetic criteria for graph drawing}
\begin{itemize}[leftmargin=1.4em]
  \item \term{Minimize edge crossings.}
  \item \term{Uniform edge length} / minimize total edge length.
  \item \term{Even / symmetric node distribution}; display symmetries.
  \item \term{Angular resolution} -- large angles between edges at a node.
  \item \term{Minimize edge bends}; few orthogonal bends.
  \item \term{Avoid node/edge overlap.}
  \item \term{Cluster separation} -- communities visually grouped.
  \item \term{Consistent flow direction} (for directed/layered graphs).
  \item \term{Preserve the mental map} across changes (dynamic graphs).
\end{itemize}

% ---------------------------------------------------------------------
\subsection{Overplotting fixes that need no preprocessing}
\begin{itemize}[leftmargin=1.4em]
  \item \term{Transparency (alpha blending)} -- density shows through.
  \item \term{Smaller marks / thinner lines.}
  \item \term{Open glyphs} (outlines instead of filled).
  \item \term{Jitter} -- small random offset to separate coincident points.
  \item \term{Subsampling on the fly} (draw a random subset).
\end{itemize}
\begin{keybox}[Contrast]
Binning/hexbin, KDE, aggregation, and dimensionality reduction \emph{do} need
preprocessing -- do not list them as ``no-preprocessing'' fixes.
\end{keybox}

% ---------------------------------------------------------------------
\subsection{Model-agnostic AI interpretability methods}
\begin{itemize}[leftmargin=1.4em]
  \item \term{Partial dependence plots (PDP)} -- average marginal effect of a
        feature.
  \item \term{Individual conditional expectation (ICE)} -- per-instance PDP
        curves.
  \item \term{Permutation feature importance} -- accuracy drop when a feature
        is shuffled.
  \item \term{LIME} -- local surrogate from input perturbations.
  \item \term{SHAP} -- Shapley-value attribution of the prediction.
  \item \term{Counterfactual explanations} -- minimal change to flip outcome.
  \item \term{Local surrogate / global surrogate models.}
  \item \term{Accumulated local effects (ALE)} -- like PDP but robust to
        correlated features.
\end{itemize}

% ---------------------------------------------------------------------
\subsection{Techniques to visualize set-typed data}
\begin{itemize}[leftmargin=1.4em]
  \item \term{Venn / Euler diagrams} -- overlapping regions; poor beyond 3--4
        sets.
  \item \term{UpSet plots} -- matrix of intersections with bar charts; scales
        to many sets.
  \item \term{Set membership matrix} -- rows = elements, columns = sets.
  \item \term{Linear / radial set diagrams} (LineSets, Kelp diagrams, BubbleSets)
        -- overlays on a spatial layout.
  \item \term{Aggregate bar charts} of set / intersection sizes.
  \item \term{Parallel sets} -- flows between categorical set memberships.
\end{itemize}

% ---------------------------------------------------------------------
\subsection{Methods for validation of visual encoding and/or interaction design}
\begin{itemize}[leftmargin=1.4em]
  \item \term{Controlled experiment / user study} -- measure time \& accuracy.
  \item \term{Qualitative usability study} / think-aloud.
  \item \term{Expert review / heuristic evaluation.}
  \item \term{Case studies} with domain experts in the field.
  \item \term{Computational / algorithmic benchmarks} -- e.g.\ quality metrics
        (stress, class consistency, scagnostics).
  \item \term{Crowdsourced experiments} (e.g.\ MTurk) for many participants.
  \item \term{Analytical evaluation} against perceptual theory
        (expressiveness/effectiveness).
  \item \term{Log analysis} of real usage; A/B tests.
\end{itemize}

% ---------------------------------------------------------------------
\subsection{Typical dependent variables in user studies}
\begin{itemize}[leftmargin=1.4em]
  \item \term{Task completion time} (speed).
  \item \term{Accuracy / error rate} (correctness).
  \item \term{Confidence} ratings.
  \item \term{Subjective preference / satisfaction.}
  \item \term{Cognitive load} (e.g.\ NASA-TLX).
  \item \term{Insights found} / recall / memorability.
  \item \term{Eye-tracking measures} (fixations, scan path).
\end{itemize}
